---
title: "Network Analysis Report"
subtitle: "`r params$outcome_of_interest`"
author: "Alessia Mapelli"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    fig_width: 10
    fig_height: 8
  pdf_document:
    toc: true
    fig_width: 10
    fig_height: 8
params:
  original_data_path: "UKB_proteo_T2D.RData"
  net_res_path: "./results/PPI_T2D_app.RData"
  outcome_of_interest: "DiabIncident"
  base_output_dir: "./results/Network_description"
  color_plots_1: "#4361ee"
  color_plots_2: "#3a0ca3"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.align = 'center',
  cache = FALSE
)

# Load required libraries
library(igraph)
library(tidygraph)
library(ggraph)
library(dplyr)
library(ggplot2)
library(backbone)
library(RColorBrewer)
library(poweRlaw)
library(knitr)
library(kableExtra)
```

```{r load-data}
# Set working directory and load data setwd(params$base_output_dir)

# Create output folder for this specific outcome
foldname <- file.path(params$base_output_dir, params$outcome_of_interest)
dir.create(foldname, recursive = TRUE, showWarnings = FALSE)

# Load network results
cat("Loading data from:", params$net_res_path, "\n")
load(params$net_res_path)
cat("Loading data from:", params$original_data_path, "\n")
load(params$original_data_path)


# Extract results
mean_res <- results$results$Cov_effect
prec_res <- results$results$Dic_adj_matrics
outcome_of_interest <- params$outcome_of_interest
prot_names <- colnames(results$additional_info$z)

cat("Analysis for outcome:", outcome_of_interest, "\n")
cat("Number of proteins:", length(prot_names), "\n")
```

# 1. Mean Effects Analysis

```{r mean-effects-analysis}
cat("Mean effects matrix dimensions:", dim(mean_res), "\n")

# Show first row of mean results
cat("First protein mean effects:\n")
print(head(mean_res[1,]))

# Check the R-squared values for the regression of each protein on covariates
covariates <- selection_data[, !(names(selection_data) %in% prot_names | names(selection_data) == "id")]
str(covariates)
 
numeric_columns <- sapply(covariates, is.numeric)
covariates[, numeric_columns] <- scale(covariates[, numeric_columns])
C <- model.matrix(~ ., covariates)
 
r2_vec <- sapply(1:length(prot_names), function(p) {
  y <- selection_data[, prot_names[p]]
  beta <- as.numeric(mean_res[p,])
  yhat <- as.numeric(as.matrix(C) %*% beta)
  ss_res <- sum((y - yhat)^2)
  ss_tot <- sum((y - mean(y))^2)
  1 - ss_res / ss_tot
})
 
cat("R-squared summary:\n")
print(summary(r2_vec))

# Identify proteins with non-zero covariate effects
cov_eff <- mean_res[, outcome_of_interest]
names(cov_eff) <- prot_names
cov_eff <- cov_eff[cov_eff != 0]
prot_cov_eff_mean <- names(cov_eff)

cat("Number of proteins with non-zero effects:", length(prot_cov_eff_mean), "\n")

if(length(prot_cov_eff_mean) > 0) {
  mean_effects_table <- data.frame(
    Protein = prot_cov_eff_mean,
    Effect = cov_eff[prot_cov_eff_mean]
  )
  
  kable(head(mean_effects_table[order(abs(mean_effects_table$Effect), decreasing = TRUE), ], 10), caption = "Top 10 Proteins with Non-Zero Mean Effects", digits = 4) %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
}
```

# 2. Network Construction and Basic Properties

```{r network-construction}
# Create network from precision matrix
outcome_matrix <- as.matrix(abs(prec_res[[outcome_of_interest]]))
cat("Matrix is symmetric:", isSymmetric(outcome_matrix, check.attributes = FALSE), "\n")

# Create igraph object
g_abs <- graph_from_adjacency_matrix(
  outcome_matrix, 
  mode = "max", 
  diag = FALSE, 
  weighted = TRUE, 
  add.colnames = NA, 
  add.rownames = NULL
)

# Basic network statistics
edge_density_val <- edge_density(g_abs, loops = FALSE)
diameter_val <- diameter(g_abs, directed = FALSE, weights = NA)

network_stats <- data.frame(
  Metric = c("Nodes", "Edges", "Edge Density", "Diameter"),
  Value = c(
    vcount(g_abs),
    ecount(g_abs),
    round(edge_density_val, 4),
    diameter_val
  )
)

kable(network_stats, caption = "Basic Network Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

cat("Nodes (V):", vcount(g_abs), "\n")
cat("Edges (E):", ecount(g_abs), "\n")
cat("Edge density:", round(edge_density_val, 4), "\n")
cat("Diameter:", diameter_val, "\n")
```

## Network Visualization

```{r network-plots, fig.width=12, fig.height=6}
# Create layout
layout <- layout_nicely(g_abs)

# Save basic network plot
plot_name <- file.path(foldname, paste0("Plots_graph_", outcome_of_interest, ".jpg"))
jpeg(plot_name, width = 485, height = 485)
plot(g_abs,
     layout = layout,
     edge.arrow.size = 1,
     vertex.color = params$color_plots_1,
     vertex.size = 10,
     vertex.frame.color = params$color_plots_1,
     vertex.label.color = "black",
     vertex.label.cex = 0.5,
     vertex.label.dist = 0,
     edge.curved = 0.2,
     edge.color = "gray",
     edge.width = 0.1,
     margin = c(0, 0, 0, 0))
dev.off()

# Plot with mean effects highlighted
vertex_colors <- rep(params$color_plots_1, vcount(g_abs))
vertex_colors[V(g_abs)$name %in% prot_cov_eff_mean] <- params$color_plots_2

plot_name <- file.path(foldname, paste0("Plots_graph_", outcome_of_interest, "_with_mean_effects.jpg"))
jpeg(plot_name, width = 485, height = 485)
plot(g_abs,
     layout = layout,
     edge.arrow.size = 1,
     vertex.color = vertex_colors,
     vertex.size = 10,
     vertex.frame.color = vertex_colors,
     vertex.label.color = "black",
     vertex.label.cex = 0.5,
     vertex.label.dist = 0,
     edge.curved = 0.2,
     edge.color = "gray",
     edge.width = 0.1,
     margin = c(0, 0, 0, 0))
dev.off()

# Display plots in report
par(mfrow = c(1, 2))
plot(g_abs,
     layout = layout,
     edge.arrow.size = 1,
     vertex.color = params$color_plots_1,
     vertex.size = 10,
     vertex.frame.color = params$color_plots_1,
     vertex.label.color = "black",
     vertex.label.cex = 0.5,
     vertex.label.dist = 0,
     edge.curved = 0.2,
     edge.color = "gray",
     edge.width = 0.1,
     margin = c(0, 0, 0, 0),
     main = "Basic Network Structure")

plot(g_abs,
     layout = layout,
     edge.arrow.size = 1,
     vertex.color = vertex_colors,
     vertex.size = 10,
     vertex.frame.color = vertex_colors,
     vertex.label.color = "black",
     vertex.label.cex = 0.5,
     vertex.label.dist = 0,
     edge.curved = 0.2,
     edge.color = "gray",
     edge.width = 0.1,
     margin = c(0, 0, 0, 0),
     main = "Network with Mean Effects Highlighted")
```

## Network Visualization - Paper Figure

```{r network-plots-paper, fig.width=12, fig.height=6}
nodes_plot_label <- match(V(g_abs)$name, prot_names)
V(g_abs)$plot_label <- nodes_plot_label

V(g_abs)$diab_influence <- ifelse(V(g_abs)$name %in% prot_cov_eff_mean, "Influenced by T2D", "Not influenced by T2D")

nod_legend <- data.frame(nodes_names = V(g_abs)$name, label= match(V(g_abs)$name, prot_names))
filename <- paste(foldname, "/Node_legend.txt", sep="")
write.table(nod_legend, file=filename)

library(scales)  
w <- E(g_abs)$weight
w[is.na(w)] <- 0
E(g_abs)$w_scaled <- rescale(w, to = c(0, 1))

plot_name <- file.path(paste0(foldname, "/Opt_plots_graph_", outcome_of_interest, "_scaled_edges_with_mean_effects.png"))

layout_figure_paper <- create_layout(g_abs, layout = 'graphopt') 

p <- ggraph(layout_figure_paper) +
  geom_edge_link(
    aes(edge_colour = w_scaled,
        edge_width = w_scaled,
        edge_alpha = w_scaled),
    show.legend = TRUE
  ) +
  
  scale_edge_colour_gradient(
    low = "white",
    high = "black",
    name = "Scaled weight"
  ) +
  
  scale_edge_width(range = c(0.2, 2.0), guide = "none") +
  scale_edge_alpha(range = c(0.25, 1), guide = "none") +
  
geom_node_point(aes(color = diab_influence), size = 4) +
  
  scale_color_manual(
    name = "Mean effect",
    values = c(
      "Not influenced by T2D" = params$color_plots_1,
      "Influenced by T2D"     = params$color_plots_2
    )
  ) +
  
  ggtitle("Protein-Protein Interaction Network") +
  theme_void() +
  theme(
    plot.margin = margin(0,0,0,0),
    plot.title = element_text(hjust = 0.5, vjust = 1),
    legend.position = "right"
  )

print(p)

ggsave(p, filename=plot_name, width=12, height=8, dpi=300)

```


## Edge Weight Distribution

```{r edge-distribution}
# Analyze edge weights
UP <- outcome_matrix[upper.tri(outcome_matrix, diag = FALSE)]
non_zero_edges <- sum(UP != 0)

edge_stats <- data.frame(
  Statistic = c("Min", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max", "Non-zero edges"),
  Value = c(summary(UP), non_zero_edges)
)

kable(edge_stats, caption = "Edge Weight Statistics", digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Plot edge weight distribution
par(mfrow = c(1, 2))
hist(UP, breaks = 100, main = "Edge Weight Distribution", xlab = "Edge Weight", col = params$color_plots_1)
plot(density(UP), main = "Edge Weight Density", xlab = "Edge Weight", col = params$color_plots_2, lwd = 2)
```

# 3. Community Detection

```{r community-detection, fig.width=10, fig.height=8}
# Community detection using fast greedy algorithm
cfg <- cluster_fast_greedy(as.undirected(g_abs))
V(g_abs)$community <- cfg$membership

num_communities <- length(unique(cfg$membership))
cat("Number of communities detected:", num_communities, "\n")

# Community size distribution
community_sizes <- table(cfg$membership)
community_stats <- data.frame(
  Community = names(community_sizes),
  Size = as.numeric(community_sizes)
)

kable(head(community_stats[order(community_stats$Size, decreasing = TRUE), ], 10), 
      caption = "Top 10 Largest Communities", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Create color palette for communities
if (num_communities <= 12) {
  colrs <- brewer.pal(max(3, num_communities), "Paired")[1:num_communities]
} else {
  # fallback for many communities
  colrs <- hue_pal()(num_communities)
}

# Save community plot
plot_name <- file.path(foldname, paste0("Plots_graph_", outcome_of_interest, "_community_det.jpg"))
jpeg(plot_name, width = 485, height = 485)
plot(g_abs,
     layout = layout,
     edge.arrow.size = 1,
     vertex.color = colrs[V(g_abs)$community],
     vertex.size = 5,
     vertex.frame.color = colrs[V(g_abs)$community],
     vertex.label = NA,
     edge.curved = 0.2,
     edge.color = "gray",
     edge.width = 0.1,
     margin = c(0, 0, 0, 0))
dev.off()

# Display community plot in report
plot(g_abs,
     layout = layout,
     edge.arrow.size = 1,
     vertex.color = colrs[V(g_abs)$community],
     vertex.size = 5,
     vertex.frame.color = colrs[V(g_abs)$community],
     vertex.label = NA,
     edge.curved = 0.2,
     edge.color = "gray",
     edge.width = 0.1,
     margin = c(0, 0, 0, 0),
     main = paste("Community Structure (", num_communities, "communities)"))
```

## Community Visualization - Paper Figure
```{r community-visualization-paper, fig.width=10, fig.height=8}
g_tbl <- as_tbl_graph(g_abs) %>%
  activate(nodes) %>%
  mutate(community = factor(V(g_abs)$community),
         node_label = V(g_abs)$plot_label)

plot_name <- file.path(paste0(foldname, "/Opt_plots_graph_", outcome_of_interest, "_community_det.jpg"))
ggsave(p, filename=plot_name, width=8, height=8, dpi=300) 

# Basic network plot
plot_name <- file.path(paste0(foldname, "/Plots_graph_", outcome_of_interest, "_community_det_weighted.png"))

p <- ggraph(g_tbl, layout = "graphopt") +
  geom_edge_link(aes(edge_colour = w_scaled, edge_width = w_scaled, edge_alpha = w_scaled),
    show.legend = TRUE) +
  scale_edge_colour_gradient(low = "white", high = "black", name = "Scaled weight") +
  geom_node_point(aes(color = community), size = 4) +
  scale_color_manual(values = colrs, name = "Community") +
  #geom_node_text(aes(label = node_label), size = 3, repel = TRUE) +
  scale_edge_width(range = c(0.2, 2.0), guide = "none") +
  scale_edge_alpha(range = c(0.25, 1), guide = "none") +
  ggtitle("Communities – Protein-Protein Interaction Network") +
  theme_void() +
  theme(plot.margin = margin(0,0,0,0),
    plot.title  = element_text(hjust = 0.5, vjust = 1))

print(p)

ggsave(p, filename=plot_name, width=12, height=8, dpi=300)

```

# 4. Node Centrality Analysis

```{r centrality-analysis}
# Calculate centrality measures
node_topology_summary <- data.frame(
  Node = V(g_abs)$name,
  cfg_community = V(g_abs)$community
)

idx <- match(node_topology_summary$Node, V(g_abs)$name)

# Calculate various centrality measures
deg <- degree(g_abs, mode = "all")
node_topology_summary$degree <- deg[idx]

eigen_centrality_result <- eigen_centrality(g_abs)
eigen_centrality_scores <- eigen_centrality_result$vector
node_topology_summary$eigen_centrality <- eigen_centrality_scores[idx]

# Handle weights for weighted measures
weights <- NULL
if (is.null(weights) && "weight" %in% edge_attr_names(g_abs)) {
  distL <- 1/E(g_abs)$weight
  weights <- E(g_abs)$weight
} else if(!is.null(weights) && any(!is.na(weights))) {
  distL <- 1/as.numeric(weights)
  weights <- as.numeric(weights)
} else {
  distL <- NA
  weights <- NA
}

node_topology_summary$page.rank <- as.numeric(page.rank(
  graph = g_abs,
  vids = V(g_abs),
  directed = FALSE,
  weights = weights
)$vector[idx])

node_topology_summary$betweenness <- as.numeric(betweenness(
  g_abs, 
  directed = FALSE, 
  weights = distL
)[idx])

node_topology_summary$hub_score <- hub_score(g_abs)$vector[idx]

# Normalize centrality measures
node_topology_summary$eigen_centrality_normalized <- scale(node_topology_summary$eigen_centrality)[, 1]
node_topology_summary$page_rank_normalized <- scale(node_topology_summary$page.rank)[, 1]
node_topology_summary$betweenness_normalized <- scale(node_topology_summary$betweenness)[, 1]
node_topology_summary$degree_normalized <- scale(node_topology_summary$degree)[, 1]
node_topology_summary$av_centrality <- rowMeans(node_topology_summary[, c(
  "degree_normalized", "eigen_centrality_normalized", 
  "page_rank_normalized", "betweenness_normalized"
)])

# Summary statistics for centrality measures
centrality_summary <- node_topology_summary %>%
  select(degree, eigen_centrality, page.rank, betweenness, hub_score, av_centrality) %>%
  summary()

cat("Average centrality summary:\n")
print(summary(node_topology_summary$av_centrality))
```

## Centrality Distributions

```{r centrality-plots, fig.width=12, fig.height=8}
# Create ggplot histograms for centrality measures
p1 <- ggplot(node_topology_summary, aes(degree)) + 
  geom_histogram(fill = params$color_plots_1, alpha = 0.7) + 
  theme_bw(12) + 
  labs(title = "Degree Distribution")

p2 <- ggplot(node_topology_summary, aes(page.rank)) + 
  geom_histogram(fill = params$color_plots_1, alpha = 0.7) + 
  theme_bw(12) + 
  labs(title = "PageRank Distribution")

p3 <- ggplot(node_topology_summary, aes(betweenness)) + 
  geom_histogram(fill = params$color_plots_1, alpha = 0.7) + 
  theme_bw(12) + 
  labs(title = "Betweenness Distribution")

p4 <- ggplot(node_topology_summary, aes(eigen_centrality)) + 
  geom_histogram(fill = params$color_plots_1, alpha = 0.7) + 
  theme_bw(12) + 
  labs(title = "Eigenvector Centrality Distribution")

# Display plots
library(gridExtra)
grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## Top Influential Nodes Analysis

```{r top-nodes-analysis}
# Identify top 10% nodes based on average centrality
threshold_centrality <- quantile(node_topology_summary$av_centrality, 0.90)
top_10_percent_nodes <- node_topology_summary[node_topology_summary$av_centrality > threshold_centrality, ]$Node
node_topology_summary$top_10_percent_central_nodes <- node_topology_summary$av_centrality > threshold_centrality

# Identify top 10% nodes based on hub score
threshold_hub <- quantile(node_topology_summary$hub_score, 0.90)
top_10_percent_nodes_hub <- node_topology_summary[node_topology_summary$hub_score > threshold_hub, ]$Node
node_topology_summary$top_10_percent_hub_nodes <- node_topology_summary$hub_score > threshold_hub

# Find intersection of top central and hub nodes
overlap_count <- sum(top_10_percent_nodes %in% top_10_percent_nodes_hub)
top_influential_nodes <- top_10_percent_nodes[top_10_percent_nodes %in% top_10_percent_nodes_hub]

cat("Top 10% central nodes:", length(top_10_percent_nodes), "\n")
cat("Top 10% hub nodes:", length(top_10_percent_nodes_hub), "\n")
cat("Overlap between central and hub nodes:", overlap_count, "\n")
cat("Top influential nodes (intersection):", length(top_influential_nodes), "\n")

# Display top influential nodes
if(length(top_influential_nodes) > 0) {
  top_nodes_info <- node_topology_summary[node_topology_summary$Node %in% top_influential_nodes, 
                                         c("Node", "degree", "eigen_centrality", "page.rank", "betweenness", "hub_score", "av_centrality")]
  
  kable(top_nodes_info, caption = "Top Influential Nodes (Central + Hub)", digits = 4, row.names = FALSE) %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
}
```

## Top Nodes Subgraph Visualization and Important Interactions retrieval

```{r top-nodes-subgraph, fig.width=10, fig.height=8}
# Create and visualize subgraph of top influential nodes
if(length(top_influential_nodes) > 0) {
  subgraph <- induced_subgraph(g_abs, vids = top_influential_nodes, impl = "copy_and_delete")
  
  subgraph_layout <- layout_nicely(subgraph)
  
  # Save top nodes plot
  plot_name <- file.path(foldname, paste0("Plots_graph_", outcome_of_interest, "_top_nodes.jpg"))
  jpeg(plot_name, width = 485, height = 485)
  plot(subgraph,
       layout = subgraph_layout,
       edge.arrow.size = E(subgraph)$weight,
       vertex.color = params$color_plots_1,
       vertex.size = 10,
       vertex.frame.color = params$color_plots_2,
       vertex.label.color = "black",
       vertex.label.cex = 0.8,
       vertex.label.dist = 0,
       edge.curved = 0.2,
       edge.color = "gray",
       edge.width = E(subgraph)$weight * 20,
       margin = c(0, 0, 0, 0))
  dev.off()
  
  # Display in report
  plot(subgraph,
       layout = subgraph_layout,
       edge.arrow.size = 1,
       vertex.color = params$color_plots_1,
       vertex.size = 10,
       vertex.frame.color = params$color_plots_2,
       vertex.label.color = "black",
       vertex.label.cex = 0.8,
       vertex.label.dist = 0,
       edge.curved = 0.2,
       edge.color = "gray",
       edge.width = 2,
       margin = c(0, 0, 0, 0),
       main = "Subgraph of Top Influential Nodes")
} else {
  cat("No top influential nodes found.\n")
}
thr <- quantile(UP, 0.99, na.rm = TRUE)
sel_edges <- E(subgraph)[E(subgraph)$weight > thr]
if (length(sel_edges) > 0) {
  e_ends <- ends(subgraph, sel_edges, names = TRUE)   # matrix: first col = "first" node
  important_df <- tibble(
      Node = e_ends[, 1],
      important_interaction = paste0(e_ends[, 1], "*", e_ends[, 2])
    ) %>%
    distinct() %>%                                 # drop accidental duplicates
    group_by(Node) %>%
    summarise(important_interaction = paste(important_interaction, collapse = "+"),
              .groups = "drop")
} else {
  important_df <- tibble(Node = character(), important_interaction = character())
}
node_topology_summary <- node_topology_summary %>%
  left_join(important_df, by = "Node")
```

## Top Nodes Subgraph Visualization - Paper Figure
```{r top-nodes-subgraph-paper-figure, fig.width=10, fig.height=8}
if(length(top_influential_nodes) > 0) {
  # Save top nodes plot
  plot_name <- file.path(foldname, paste0("Opt_plots_graph_", outcome_of_interest, "_top_nodes_with_weigths.jpg"))
  
  g_tbl <- as_tbl_graph(subgraph) %>%
  activate(nodes) %>%
  mutate(community = factor(V(subgraph)$community),
         node_label = V(subgraph)$plot_label,
         node_name = V(subgraph)$name)
  
  layout <- create_layout(g_tbl, layout = 'graphopt') 
  p <- ggraph(layout) +
    geom_edge_link(aes(edge_colour = w_scaled, edge_width = w_scaled, edge_alpha = w_scaled),
    show.legend = TRUE) +
  scale_edge_colour_gradient(low = "white", high = "black", name = "Scaled weight") +
    geom_node_point(color = params$color_plots_1, size = 4) +
    geom_node_text(aes(label = node_name), size = 4, repel = TRUE) +
      scale_edge_width(range = c(0.2, 2.0), guide = "none") +
  scale_edge_alpha(range = c(0.25, 1), guide = "none") +
    ggtitle("Top nodes - Protein-Protein Interaction Network") +
    theme_void() +
    theme(plot.margin = margin(0,0,0,0),
      plot.title  = element_text(hjust = 0.5, vjust = 1))

  print(p)
  
  ggsave(p, filename=plot_name, width=12, height=8, dpi=300)
  
  cat("Subgraph nodes:", vcount(subgraph), "\n")
  cat("Subgraph edges:", ecount(subgraph), "\n")
}
```

# 5. Summary and Export

```{r summary-export}
# Save node topology summaries
file_name_original <- file.path(foldname, paste0("Summary_node_topology_", outcome_of_interest, ".csv"))
write.csv(node_topology_summary, file_name_original, row.names = FALSE)

cat("Node topology summaries saved:\n")
cat("  Original network:", file_name_original, "\n")
```

**Generated on:** `r Sys.time()`

**Analysis Parameters:**
- Data file: `r params$net_res_path`
- Outcome: `r params$outcome_of_interest`
- Output directory: `r foldname`
- Colors: Primary = `r params$color_plots_1`, Secondary = `r params$color_plots_2`
